---
title: "IST686 Problem Set 1 Spring 2024"
author: "Michael A. d'Amore"
date: "2024-09-01"
output: pdf_document
---

```{r }
install.packages("LaTex")


```

The homework for week one is exercises 1, 3, and 4 on page 20. 

Attribution statement: (choose only one and complete as necessary)
1. I did this homework by myself, with help from the book and the professor. 
2. I did this homework with help from the book and the professor and these Internet sources:
3. I did this homework with help from <Name of another student> but did not cut and paste any code.

# Chapter 1, Exercise 1

_Using the material from this chapter and possibly other information that you look up, write a brief definition of these terms in your own words: mean (aka average), median, mode, variance, standard deviation, histogram, normal distribution, and Poisson distribution. (1 point for each definition)_

* Mean: "Average"

* Median: The middle point of a set of numbers

* Mode: Number that appears the most

* Variance: How much the number differs from the average. Its the average of squared differences. Think about is as a plus minus.    

* Standard deviation: Its a number that tells you, on average, how far the numbers in your set are from the mean.

* Histogram: A bar chart that divided into bins. Those bins have a range and you plot how many times that range occurs 

* Normal distribution: A type of graph used to describe how data points are spread out around the mean (average). Usually the shape is a bell curve.

* Poisson distribution: Poisson distribution helps us understand and predict the likelihood of a certain number of events happening over a fixed period or in a given space. The Poisson dist is used to predict the likelihood of a certain event happening in a fixed time. There's no upper limit. We can take Chat GPT's example,[if you usually get about 10 emails per hour, the Poisson distribution can tell you the probability of getting exactly 5 emails in the next hour].


# Chapter 1, Exercise 3

_Use the data() function to get a list of the data sets that are included with the basic installation of R: just type “data()” at the command line and press enter._ 

```{r }
data()

# Load the dataset
data(euro)

# Get a summary
summary(euro)

#Min: This is the smallest value in the dataset. It represents the lowest exchange rate recorded in the dataset.
#1Q: This is the value below which 25% of the data fall. It marks the lower quarter of the dataset.
#Median: The median is the middle value of the dataset when it is ordered from lowest to highest. This means that 50% of the exchange rates are below 13.7603, and 50% are above. 
#Mean: average
#3Q: This is the value below which 75% of the data fall. It marks the upper quarter of the dataset.
#Max: Highest value in the dataset 
```

_Choose a data set from the list that contains at least one numeric variable--for example, the Biochemical Oxygen Demand (BOD) data set. Use the summary() command to summarize the variables in the data set you selected--for example, summary(BOD). (Pick something other than BOD though.) (1 pt) Write a brief description of the mean and median of each numeric variable in the data set. (1 pt for each value) Make sure that in presenting the values, you define what a “mean” and a “median” are, that is, the technical definition and practical meaning of each of these quantities. (1 pt for each definition)_

# Chapter 1, Exercise 4

_As in the previous exercise, use the data() function to get a list of the data sets that are included with the basic installation of R. Choose a data set and pick out one variable, for example, the LakeHuron data set (levels of Lake Huron in the years 1875 through 1972). Use the hist() command to create a histogram of the variable--for example, hist(LakeHuron). (Again, pick something other than LakeHuron.) (2 pts) Describe the shape of the histogram in words (e.g., spread out or compact, symmetric or skewed, peaked or flat, etc.). (2 pts)  Which of the distribution types we talked about in class or that you've read about do you think these data fit most closely (e.g., normal, Poisson). (2 pts) Speculate on why your selected data may fit that distribution, i.e., in what way does the phenomenon (e.g., lake levels) match the processes we talked about in class that leads to the data being distributed as they are? (2 pts)_ 

```{r }
# Load the built-in data
data()

# Load the dataset
data(AirPassengers)

#histogram data
hist(AirPassengers, main="Histogram of AirPassengers", xlab="Number of Passengers", ylab="Frequency", col="lightblue")

#Spread: The histogram shows that the data is spread out, with most of the passenger counts falling between 100 and 500. The counts are more frequent at the lower end (between 100 and 200) and gradually decrease as you move toward higher passenger counts.

#Symmetry: The histogram is right-skewed (positively skewed). There are more data points on the lower end and fewer data points as you move to higher numbers.

#Peakedness: The histogram has multiple peaks, indicating that certain ranges of passenger counts are more common. The most frequent values seem to be around 100 to 200 passengers.

#It looks like a Poisson but it also has a log-normal shape it could be Poisson as well because we are measuring random frequency of an event occurring.

#The right-skewed distribution makes sense for this dataset because the number of airline passengers generally increases over time, but the increase isn't uniform. Early on, there were fewer passengers as air travel was less common, and as it became more accessible and popular.

```
# Additional exercises 

Exercises 1.1 through 1.4 and 3.1 through 3.5 from Akinkunmi, M. (2019). Introduction to Statistics Using R. Springer Cham. (https://www.amazon.com/Introduction-Statistics-Synthesis-Lectures-Mathematics/dp/1681735091)




