
---
title: "IST772 Problem Set 4"
author: "Michael A. d'Amore"
output:
  pdf_document: default
---

The homework for week 4 is based is based on exercises 7-10 on page 66, but with changes as noted in the text in this notebook (i.e., follow the problems as given in this document and not the textbook). 

Attribution statement: (choose only one)
1. I did this homework by myself, with help from the book and the professor

```{r}

wine <- read.csv(file.choose())

```


# Chapter 4, Exercise 7 

_The wine data set available on Blackboard contains data for properties of different wines (see https://archive.ics.uci.edu/dataset/109/wine for documentation). The wines are labelled in the class variable as 1, 2 or 3. Run the summary() command on the dataset and explain the output. Create a histogram of the hue content for wine 1 (1 pt). As a reminder about R syntax, here is one way that you can access the class 1 hue data:_

```{r eval=FALSE}

library(ggplot2)

wine$hue[wine$class==1]
summary(wine)
class_1_wines <- wine[wine$class == 1, ]

ggplot(class_1_wines, aes(x = hue)) +
  geom_histogram(binwidth = 0.1, fill = "lightblue", color = "black") +
  labs(title = "Histogram of Hue for Class 1 Wines", x = "Hue", y = "Frequency") +
  theme_minimal()

```

Using the [dplyr](https://dplyr.tidyverse.org/) package, you can instead write:

```{r}
library(dplyr)

wine %>% filter(class == 1) %>% select(hue)
```

(Note that a select function is defined in multiple packages, so if you want to be sure you're using the one from the dplyr library, call dplyr::select.)

_Also create histograms of the hue level for wines 2 and 3. What can you say about the differences in the hue level by looking at the histograms? (1 pt)_

```{r}
library(ggplot2)
class_2_wines <- wine[wine$class == 2, ]
class_3_wines <- wine[wine$class == 3, ]

ggplot(class_2_wines, aes(x = hue)) +
  geom_histogram(binwidth = 0.1, fill = "lightblue", color = "black") +
  labs(title = "Histogram of Hue for Class 2 Wines", x = "Hue", y = "Frequency") +
  theme_minimal()

ggplot(class_3_wines, aes(x = hue)) +
  geom_histogram(binwidth = 0.1, fill = "lightblue", color = "black") +
  labs(title = "Histogram of Hue for Class 3 Wines", x = "Hue", y = "Frequency") +
  theme_minimal()


#Class 1 and 2 wines have hue values that are more spread out and range into higher hue values, 
#while Class 3 wines are more concentrated around lower hue values.

#Class 3 wines have a more skewed distribution, concentrated in the lower end of the hue spectrum, which suggests #these wines may have a consistently lower hue compared to the other classes.
#Class 2 wines have a wider range of hue values and appear more variable compared to class 1 and class 3 wines.

#My interpretation is that class 1 corresponds to red wine lower hue color, darker colors absorb more light
#The broader hue distribution class 2 (peaking around 0.9 but extending to 1.8) suggests these wines might sit between red and white in color, which is characteristic of rosé.
#Class 3 is by elimination white but also they have a higher hue color corresponding to the lighte color wine
```

# Chapter 4, Exercise 8

_Create a boxplot (or violin plot) of the hue data, using the model “hue ~ class" so they are plotted side-by-side. (1 pt) What can you say about the differences in the hue by looking at the boxplots for the different wines? (1 pt)_


```{r}
library(ggplot2)
ggplot(wine, aes(x = factor(class), y = hue)) +
  geom_violin(trim = FALSE, fill = "lightblue") +
  labs(title = "Violin Plot of Hue by Wine Class", x = "Wine Class", y = "Hue") +
  theme_minimal()

#As I stated in my previous answer my hypothesis is further proved how 2 classes are simmilar and their qualities #while the 3rd is completeky different,

#Class 1 wines have a more consistent hue, representing white wines with light and consistent coloring.
#Class 2 wines have a broader hue range, which indicates rosé wines or other wines with intermediate hue values.
#Class 3 wines have lower hue values, these are red wines, known for having deeper and darker hues.

```

# Chapter 4, Exercise 9

_Run a t‐test to compare the means of classes 1 and 2 in the wine data and report the confidence interval. (1 pt) Give an interpretation of the confidence interval. (1 pt) Make sure to include a carefully worded statement about what the confidence interval implies with respect to the population mean difference between the wines (specifically, if the groups are different or not). (1 pt)_

```{r}
t_test_result <- t.test(hue ~ class, data = wine, subset = class %in% c(1, 2))

# Display the results of the t-test
print(t_test_result)


#The t-test results show no statistically significant difference in the mean hue between class 1 and class 2 #wines (p-value = 0.8402). The 95% confidence interval for the difference in means is [-0.0506, 0.0621], which #includes 0. This indicates that the hue levels of the two wine classes are likely similar, and there is no #evidence to suggest a meaningful difference between them in terms of hue.

#Although my theory still stands and my new hypothesis is the wines used in the class 2 are really light colored #roses such as pinot noir that are very close to white instead of a darker rose such as Merlot or Cabernet
```


# Chapter 4, Exercise 10

_Run a t‐test to compare the means of wines 1 and 3 in the wine data. (1 pt) Report and interpret the confidence interval. (1 pt + 1 pt for statement about means)_
```{r}
# Perform a t-test between class 1 and class 3 wines for hue
t_test_1_3 <- t.test(hue ~ class, data = wine, subset = class %in% c(1, 3))

# Print the t-test results
print(t_test_1_3)

#The 95% confidence interval for the difference in means is [0.3348, 0.4238], indicating that the mean hue of #class 1 wines is significantly higher than that of class 3 wines. 
#Here theres a big difference between hues 1 and 3 my theory stands about the type of hue colors of the wine. 
```

